# Introduction to Bayesian hypothesis testing

In this chapter, we will introduce an alternative to the null-hypothesis significance testing procedure employed up to now. 

## Fundamentals of Bayesian inference

### Probability in times of Covid

Let's start with a classic example in a topical guise. As we hopefully move to a new phase in the covid-19 pandemic,^[I'm writing this on 3 December 2020] the UK government plans to employ mass testing as part of a wider strategy also involving vaccination ([Guardian, 1 December 2020](https://www.theguardian.com/world/2020/dec/01/funds-for-mass-covid-testing-offered-to-local-authorities-in-tier-3)). The testing will involve lateral flow tests, which are relatively inexpensive and give a result in about 20 minutes. Estimates are that these tests give a positive test result in 76.8% of cases of covid-19, and a negative test result in 99.68% of cases ([Department of Health and Social Care, 11 November 2020](https://www.gov.uk/government/news/oxford-university-and-phe-confirm-high-sensitivity-of-lateral-flow-tests)). The true positive rate (76.8%) is also called the __sensitivity__ of a test, and the true negative rate (99.68%) the __specificity__. of On 26 November, the [Office of National Statistics ](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/conditionsanddiseases/bulletins/coronaviruscovid19infectionsurveypilot/26november2020) estimates the rate of covid-19 cases in the general population of England to be 1.17%. That implies that about 1 in 85 people carry the virus, which is also called the __base rate__. Suppose someone is tested and the test result is positive. What is the probability that they are currently infected with covid-19? The probability is _not_ .768. This is the conditional probability $P(\text{positive test}|\text{covid-19})$. What we would like to know however is a different conditional probability, namely $P(\text{covid-19}|\text{positive test})$. And these are _not_ the same!

We can work out the desired probability using the rules of probability discussed in Section \@ref(sec:02-rules-of-probability). 
$$P(\text{covid-19}|\text{positive test}) = \frac{P(\text{covid-19 and positive test})}{P(\text{positive test})}$$
From the multiplication rule, we know that
$$P(\text{covid-19 and positive test}) = P(\text{positive test}|\text{covid-19}) \times P(\text{covid-19})$$ 
To work out $P(\text{positive test})$, we need to consider all the ways in which someone can obtain a positive test result. In this situation, there are two: the person can carry covid-19 and have a positive test result, or the person can not carry covid-19 and obtain a positive test result. Thus 
$$P(\text{positive test}) = P(\text{covid-19 and positive test}) + P(\text{no covid-19 and positive test})$$
We have already specified how to calculate $P(\text{covid-19 and positive test})$. Similarly, we can compute $P(\text{no covid-19 and positive test})$ as
$$P(\text{no covid-19 and positive test}) = P(\text{positive test}|\text{no covid-19}) \times P(\text{no covid-19})$$
Now we are in a position to calculate $P(\text{covid-19}|\text{positive test})$ from the sensitivity and specificity of the test, and the base rate of covid-19 infection. We know that $P(\text{covid-19}) = .0117$, hence $P(\text{no covid-19}) = 1 - P(\text{covid-19}) = 1 - .0117 = .9883$. Putting all the numbers in a table:
```{r covid-mass-test-sensitivy-specificty}
tab <- data.frame(c("76.8%","23.2%"),c("0.32%","99.68%"))
rownames(tab) <- c("positive test","negative test")
knitr::kable(tab, col.names = c("covid-19 (1.17%)","no covid-19 (98.83%)"), align=c("r","r"))
```

$$\begin{align}
P(\text{covid-19}|\text{positive test}) &= \frac{P(\text{covid-19 and positive test})}{P(\text{positive test})} \\
&= \frac{P(\text{positive test}|\text{covid-19}) P(\text{covid-19})}{P(\text{pos. test}|\text{covid-19}) P(\text{covid-19}) + P(\text{pos. test}|\text{no covid-19}) P(\text{no covid-19})} \\
&= \frac{.768 \times .0117}{.768 \times .0117 + .0032 \times .9883} \\
&= 0.7396
\end{align}$$
So just under 3 out of all 4 people that positive in this scenario would actually carry covid-19. As a result, 1 out of 4 people might be asked to quarantine without really needing to do so. Although of course very unfortunate for those people, that does not seem like a too-high price to pay to me. But the base-rate is very important here. If the rate of covid-19 infections is lowered to $P(\text{covid-19}) = .01$ (i.e. 1%), then the result would be $P(\text{covid-19}|\text{positive test}) = 0.1937$, which means that only about one in five people who test positive are actually infected by covid-19! When the base-rate is lowered, massive testing seems like a much less useful procedure.

Perhaps the equations seem a little abstract. Another way to explain the resulting conditional probability is through the tree diagram of Figure \@ref(fig:covid-19-test-tree). The tree represents a group of 100,000 people from the general population, of which 1,176 would have covid-19, and 903 of these would also get a positive test result. Of the 98,824 people without covid-19, 316 would receive a positive test result. While the change of a false positive is very low, because so many people do not have covid-19, the actual frequency of people without covid-19 who obtain a positive test result is not that much smaller than the number of people with covid-19 who obtain a positive test result. The conditional probability can then be computed simply as 
$$P(\text{covid-19}|\text{positive test}) = \frac{903}{903 + 316} = .74$$
which is equal to the value computed earlier (up to rounding error resulting from converting probabilities to whole people).

```{r covid-19-test-tree, fig.cap="Outcome tree representing mass testing for covid-19."}
library(DiagrammeR)

treePlot <- "
  digraph cvid_tree_plot {

    graph [layout = dot,
           rankdir = LR]

    node [shape = box] 
    S [label = '100,000 people']
    
    node [shape = oval, style = filled]
    H [label = 'covid-19', fillcolor = '#cc0000']
    T [label = 'no covid-19', fillcolor = '#00cc00']
    HH [label = 'positive test', fillcolor = '#ff0000' ]
    HT [label = 'negative test', fillcolor = '#990000']
    TH [label = 'positive-test', fillcolor =  '#00ff00']
    TT [label = 'negative-test', fillcolor = '#009900']
    
    edge [label = '1,176']
    S -> {H}
    
    edge [label = '98,824', labelloc=b]
    S -> {T}
    
    edge [label = '903']
    H -> {HH}
    
    edge [label = '273', labelloc=b]
    H -> {HT}
    
    edge [label = '316' ]
    T -> {TH}
    
    edge [label = '98,508', labelloc=b]
    T -> {TT}
}
"
DiagrammeR::grViz(treePlot)
```

## Bayes' rule

In calculating a conditional probability from other conditional probabilities and base-rates, we have just applied the general rules of probability. That's nothing special, really. In abstract notation, the formula known as Bayes' rule is

\begin{equation}
P(A|B) = \frac{P(B|A) P(A)}{P(B|A)P(A) + P(B|\neg A) P(\neg A)}
(\#eq:bayes-rule)
\end{equation}

Again, there is nothing special about this formula itself, it follows from the rules of probability. These rules were however not clearly specified when Reverend Thomas Bayes defined the rule in an essay which was posthumously published [@bayes1763]. More importantly, he used the rule to infer an unknown parameter of a statistical model. According to the Frequentist View (see Section \@ref(sec:02-probability-definition)), a parameter has a true value, but you cannot assign a probability to it, because it is not a random event that has a long-run frequency. It just has one value: the true value.

In assigning probabilities to parameters, Thomas Bayes can be seen as the founding father of the Subjectivist View of probability. There has been quite a lot of philosophical discussion about probability interpretations. The subjectivist view is that a probability represents a rational degree of belief. This belief can be about anything, whether actual events in the world, or more abstract concepts such as hypotheses or model parameters. Bayesian inference concerns adjusting prior beliefs in light of evidence. The resulting adjusted belief is called the posterior belief. In the previous example, the base-rate of covid-19 infections can be seen as a rational prior belief that a randomly chosen person from the general population in England has covid-19. Upon observing a positive test result, this __prior probability__ $P(\text{covid-19})$ can be adjusted to become the __posterior probability__  $P(\text{covid-19}|\text{positive test})$.

### We missed you Paul!

In Bayesian statistics, we can apply the principles of Bayesian inference to anything we can assign degrees of belief to. For instance, our belief that Paul the Octopus had psychic abilities. In our general model of Paul's predictions (Section \@ref(sec:02-binomial-model), Equation \@ref(eq:definition-binomial-distribution)), we assumed there was a probability that he made a correct prediction, which we denoted by $\theta$. This parameter probability could in principle take any value $0 \leq \theta \leq 1$. The idea of a prior distribution for such a parameter is to assign to each possible value of $\theta$ a "degree of belief" that this is the true value. These degrees of belief should obey the rules of probability. In the coin-flipping model, which assumed Paul was randomly guessing, there was only one possible value, namely $\theta=.5$. That means that, if we were to believe this model is true, we would consequently  believe that any other value is impossible: $P(\theta \neq .5) = 0$, which implies $P(\theta = .5) = 1$. If we don't believe that Paul is necessarily randomly guessing, then the parameter could have other values as well. Figure \@ref(fig:prior-distributions-Paul) shows two possible prior distributions. In the plot on the left, the prior assigns an equal probability to any possible value of $\theta$. This is also called a uniform distribution, and reflects the beliefs of someone who considers that "anything goes" when it comes to Paul's ability to predict the outcome of football matches. In the plot on the right, the prior distribution reflects the beliefs of someone who quite strongly considers Paul a good predictor of the outcome of football matches.  

```{r prior-distributions-Paul, fig.cap="Two different prior distributions for the probability that Paul makes a correct prediction, the normalised likelihood function ($P(Y|\\theta)/P(Y)$) and the resulting posterior distributions after observing that Paul made $Y=12$ out of $n=14$ correct predictions. Wherever the normalised likelihood is larger than 1, the posterior probabiilty is larger than the prior probability, while the posterior probability is lower than the prior probability wherever the normalised likelihood is lower than 1. Note that in MODEL 1, the normalised likelihood and posterior distribution are identical and therefore overlapping on the plot.", fig.width=6,fig.height=3.5, out.width="90%"}
library(dplyr)
library(ggplot2)

data.frame(model=rep(c("MODEL 1","MODEL 2"),each=3*100),type=rep(c("prior","likelihood","posterior"),each=100),x=seq(0,1,length=100)) %>%
  mutate(type=factor(type,levels=c("prior","likelihood","posterior")), y=if_else(model == "MODEL 1",case_when(type=="prior" ~ dbeta(x,1,1), type=="likelihood" ~ dbinom(12,14,x)/(choose(14,12)*(beta(1+12,1+2)/beta(1,1))), type == "posterior" ~ dbeta(x,1+12,1+2)),case_when(type=="prior" ~ dbeta(x,12,2), type=="likelihood" ~ dbinom(12,14,x)/(choose(14,12)*(beta(12+12,2+2)/beta(12,2))), type == "posterior" ~ dbeta(x,12+12,2+2)))) %>%
  ggplot(aes(x=x,y=y,colour=type, group=type)) + geom_line(lwd=1.5,alpha=.5) + facet_wrap(~model) + xlab(expression(theta)) + ylab(expression(p(theta))) + theme(legend.position="bottom", legend.title = element_blank())
```

We can write the posterior distribution for the parameter $\theta$, conditional upon observed data $Y$ (e.g., $Y=12$ out of $n=14$ correct predictions) as:
\begin{align}
p(\theta|Y) &= \frac{P(Y|\theta) \times p(\theta)}{P(Y)} \\
&= \frac{P(Y|\theta)}{P(Y)} \times p(\theta)
(\#eq:posterior-parameter-specification)
\end{align}
Here, $P(Y|\theta)$ is the likelihood function (the probability of the observed data $Y$ given a particular value of $\theta$), $p(\theta)$ is the prior distribution of the parameter $\theta$, and $P(Y)$ is the probability of the data over all possible values of $\theta$ (weighted by their prior probability). This is also called the __marginal likelihood__, and technically is defined as
$$P(Y) = \int P(Y|\theta) p(\theta) d \theta$$
i.e. as the integral of the product of the likelihood function and prior over all possible values of $\theta$, but you don't need to worry about this. The formulation on the second line of Equation \@ref(eq:posterior-parameter-specification) is just a rearrangement of the terms, but is meant to show that you can think of the posterior probability as the product of the prior probability and the __normalised likelihood__ (the likelihood of the data for a particular value of $\theta$ compared to the marginal likelihood of $Y$ over all possible values of $\theta$).  

In words, we can describe Equation \@ref(eq:posterior-parameter-specification) as:
$$\begin{align}
\text{posterior} &= \frac{\text{likelihood} \times \text{prior}}{\text{marginal likelihood}} \\
&= \text{normalised likelihood} \times \text{prior}
\end{align}$$
The values of the normalized likelihood and the resulting posterior distribution are also shown in Figure \@ref(fig:prior-distributions-Paul). Because the posterior probabilities are calculated by multiplying the prior probability by the normalised likelihood, the posterior probability will be _higher_ than than the prior probability when the normalised likelihood is _larger_ than 1. You can think of the normalised likelihood as an average likelihood over all possible values of $\theta$ (we will discuss this in a little more detail soon). So when a particular value of $\theta$ assigns a higher likelihood to the data than the average likelihood of the data, i.e. $\frac{P(Y|\theta)}{P(Y)} > 1$, the evidence that this value of $\theta$ is the true value increases. Conversely, the posterior probability will be _lower_ than than the prior probability when the normalised likelihood is _smaller_ than 1. So when a particular value of $\theta$ assigns a lower likelihood to the data than average, the evidence that this value of $\theta$ is the true value decreases. This, in a nutshell, is how Bayesian inference of parameters works.

Inspecting the posterior distributions resulting from the two models in Figure \@ref(fig:prior-distributions-Paul), you can see that the posterior distribution is mostly located at the higher values of $\theta$, because small values of $\theta$ are not very likely given Paul's success rate of 12 out of 14 correct predictions. Comparing the posterior distributions between the two models, you may also see that the posterior distribution for MODEL 2 is less dispersed (i.e. "narrower") than for MODEL 1. This is a direct consequence of the difference in the prior distributions. The uniform distribution of MODEL 1 does not make very precise predictions regarding the likely values of $\theta$ ("anything goes"). The more dispersed the prior is, the more dispersed the posterior will be. The prior distribution of MODEL 2 is, in some sense, much more daring, indicating that low values of $\theta$ are very unlikely _a priori_. Because these values are not believed to be true in the first place, they are also relatively less likely after observing the data. 

### The marginal likelihood and prior predictive distribution

A Bayesian statistical model consists of both the prior distribution and the likelihood function. Both are integral parts of a Bayesian model. This is different from statistical models in the Frequentist tradition, which only focus on the likelihood. One way to view a Bayesian model is as a _hierarchical model_, similar how linear mixed-effects models can be viewed as hierarchical models. With this formulation, it is straightforward to simulate a Bayesian model, by first sampling a random parameter value from the prior distribution $p(\theta)$, and then using this to sample a value of the dependent variable from the conditional distribution $P(Y|\theta)$. The resulting marginal distribution of the dependent variable, $P(Y)$, is also called the __prior predictive__ distribution. The sampling scheme is
\begin{align}
Y_j &\sim P(Y|\theta_j) \\
\theta_j &\sim p(\theta)
\end{align}
Not only can we use each sampled parameter value $\theta_j$ to sample a value of the dependent variable, we can also use each to compute the likelihood of the data actually observed, e.g. $P(Y=12|\theta_j)$. Figure \@ref(fig:prior-predictive-samples-binomial-bayesian-models) shows 1000 samples parameter values (prior samples), corresponding samples of the dependent variable (prior predictive samples), and the likelihood value for Paul's predictions (sampled likelihood), for both models in Figure \@ref(fig:prior-distributions-Paul). Note that all these are based on the same set of sampled parameter values $\theta$. 

```{r prior-predictive-samples-binomial-bayesian-models, fig.width=6,fig.height=3.5, fig.cap="Number of correct predictions (out of 14) for 1000 simulations from both Bayesian models of Figure \\@ref(fig:prior-distributions-Paul).", fig.show='hold'}
set.seed(91237898)
tdat <- rbind(
  data.frame(model = "MODEL 1", theta = rbeta(1000,1,1)),
  data.frame(model = "MODEL 2", theta = rbeta(1000,12,2))
) %>% mutate(likelihood=dbinom(12,14,prob=theta))
tdat$Y = rbinom(2000,size=14,prob=tdat$theta)
tdat %>%
  ggplot(aes(x=theta)) + geom_histogram(colour="black", binwidth=.025) + facet_wrap(~model) + ggtitle("Prior samples") + xlab(expression(theta))

tdat %>%
  ggplot(aes(x=Y)) + geom_histogram(colour="black", binwidth=1) + facet_wrap(~model) + ggtitle("Prior predictive samples")

tdat %>%
  ggplot(aes(x=likelihood)) + geom_histogram(colour="black", binwidth=.01) + facet_wrap(~model) + ggtitle("Sampled likelihood") + xlab("Likelihood (k=12, n=14)")

```

You can see that the prior predictive distribution looks quite similar in shape to the prior distribution, which is because for this model, there is a close link between the probability of a correct prediction ($\theta$), and the total number of correct predictions ($Y$). You can also see that the likelihood value is relatively low for many sampled parameters from MODEL 1. That is because there are many relatively low sampled values of $\theta$ in MODEL 1, for which the likelihood of 12 out of 14 correct predictions is low. Model 2 does better in accounting for the observed data, with generally higher likelihood values for the sampled parameters. This better fit to the observed data can also be seen in the prior predictive distribution of MODEL 2, where 12 correct predictions is a common occurrence. In MODEL 1, all numbers of correct predictions are sampled roughly equally often.

The sampled likelihoods can be used to compute an average likelihood for each model. This average is an estimate of the marginal likelihood, and computed simply as
$$\hat{P}(Y=12) = \sum_{j=1}^n P(Y|\theta_j)$$
Because most of the sampled likelihoods are low for MODEL 1, the average is relatively low as well: $\hat{P}(Y=12) = `r format(mean(tdat$likelihood[tdat$model == "MODEL 1"]),digits=3)`$. For MODEL 3, most of the sampled likelihoods are relatively high, and hence the average is higher than for MODEL 1:  $\hat{P}(Y=12) = `r format(mean(tdat$likelihood[tdat$model == "MODEL 2"]),digits=3)`$. Because MODEL 2 on average assigns a higher probability to the observed data, it offers a better account of the observed data than MODEL 1. As we will see, a Bayesian version of the likelihood ratio called the Bayes Factor, which reflects the ratio of marginal likelihoods for models, is central in the Bayesian version of hypothesis testing.

Before we go on to look at the Bayes Factor in more detail, I would like to point out a close correspondence between the marginal likelihood and the prior predictive distribution. In fact, the marginal likelihood is just the prior predictive distribution evaluated at the observed data point. The prior predictive distribution is the marginal distribution of the data, whilst the marginal likelihood is the value of this distribution for a particular value of the dependent variable. So we can also estimate the marginal likelihood by computing the relative frequency of $Y=12$ in the prior predictive distribution. For MODEL 1, this gives $\hat{P}(Y=12) = `r sum(tdat$Y[tdat$model == "MODEL 1"] == 12)/1000`$, and for MODEL 2, it is $\hat{P}(Y=12) = `r sum(tdat$Y[tdat$model == "MODEL 2"] == 12)/1000`$. These two ways of estimating the marginal likelihood are not exactly the same, because they approach the same quantity via a different route.^[The first method is generally more reliable.] And both are subject to random sampling variation due to simulating a limited number of parameter values. Accurately computing marginal likelihoods is, for most models, a rather complicated thing, but we won't go into the details here. 

### Hypothesis testing, relative evidence, and the Bayes Factor

In the Frequentist null-hypothesis significance testing procedure, we defined a hypothesis test in terms of comparing two nested models, a general MODEL G and a restricted MODEL R which is a special case of MODEL G. Moreover, we defined the testing procedure in terms of determining the probability of a test result, or one more extreme, given that the simpler MODEL R is the true model. This was necessary because MODEL G is too vague to determine the sampling distribution of the test statistic.

By supplying a prior distribution to parameters, Bayesian models can be "vague" whilst not suffering from the problem that they effectively make no prediction. As we saw in the prior predictive distributions in Figure \@(fig:prior-predictive-samples-binomial-bayesian-models), even MODEL 1, which assumes all possible values of the parameter $\theta$ are equally likely, still provides a valid predicted distribution of the data. Because any Bayesian model with a valid prior distribution provides a valid prior predictive distribution, which then also provides a valid value for the marginal likelihood, we do not have to worry about complications that arise when comparing models in the Frequentist tradition, such as that the likelihood of one model will always be higher than the other because we need to estimate an additional parameter by maximum likelihood. The relative marginal likelihood of the data assigned by each model, which can be stated as a marginal likelihood ratio analogous to the likelihood ratio of Chapter \@ref(ch:modeling), provides a direct measure of the relative evidence for both models. The marginal likelihood ratio is also called the Bayes Factor, and can be defined for two general Bayesian models as:
\begin{equation}
\text{BayesFactor}(\text{MODEL 1}, \text{MODEL 2}) = \frac{P(Y_1,\ldots,Y_n|\text{MODEL 1})}{P(Y_1,\ldots,Y_n|\text{MODEL 2})}
(\#eq:Bayesfactor-definition)
\end{equation}
where $P(Y_1,\ldots,Y_n|\text{MODEL } j)$ denotes the marginal likelihood of observed data $Y_1,\ldots,Y_n$ according to MODEL $j$. 


When, upon observing one event (e.g., a positive test result) the posterior probability of another event (e.g., covid-19) increases, that means that the posterior belief of the complementary or opposite event (e.g., not covid-19) will decrease. 
$$\begin{align}
\frac{P(A|B)}{P(\neg A|B)} &= \frac{\frac{P(B|A) \times P(A)}{P(B|A) \times P(A) + P(B|\neg A) \times P(\neg A)}{\frac{P(B|\neg A) \times P(A)}{P(B|A) \times P(A) + P(B|\neg A) \times P(\neg A)} \\

\end[align}$$


\\
&= \frac{P(E|H_0)}{P(E|H_1)} \times \frac{P(H_0)}{P(H_1)} \\
\text{Posterior odds} &= \text{Bayes factor} \times \text{Prior odds}
\end{align}


The relative impact of the first event on the belief in the second event and its complement is sometimes also called the __evidence__. 

\begin{align}
\frac{P(A|B)}{P(\neg A|B)} &= \frac{\frac{P(E|H_0) \times P(H_0)}{P(E)}}{\frac{P(E|H_1) \times P(H_1)}{P(E)}} \\
&= \frac{P(E|H_0)}{P(E|H_1)} \times \frac{P(H_0)}{P(H_1)} \\
\text{Posterior odds} &= \text{Bayes factor} \times \text{Prior odds}
\end{align}

## A Bayesian t-test


## Common objections to null-hypothesis significance testing

### The $p$-value is not a proper measure of evidential support

For a given alternative to $H_0$, the (expected) $p$-value becomes smaller and smaller as the sample size increases. But smaller and smaller effect sizes become significant when sample size increases. Under $H_0$, any $p$-value is equally likely (the distribution of $p$-values is uniform)

```{r,fig.width=6,fig.height=4, fig.show='hold'}
f <- seq(0,6,length=100)
plot(f,df(f,df1=1,df2=95),type="l",xlab="F",ylab="density")
tf <- seq(2.829,7,length=1000)
polygon(x=c(tf[1],tf,tf[length(tf)]),y=c(0,df(tf,df1=1,df2=95),0),density=NA, col="#0080ff")
#lines(x=c(tf[1],tf[1]),y=c(0,.5),lty=2)
#text(tf[1],.6,labels=expression(paste("critical value (",alpha==.05,")")))
lines(x=c(2.829,2.829),y=c(0,1),lty=2)
text(2.829,1.1,labels="actual value")

xx <- seq(0,1,length=10)
plot(xx,rep(1,length(xx)),type="l",xlab="p-value",ylab="density")
polygon(x=c(1-pf(tf[1],df1=1,df2=95),1-pf(tf[1],df1=1,df2=95),0,0),y=c(0,1,1,0),density=NA, col="#0080ff")
```

### The $p$-value depends on researcher intentions

The sampling distribution of a test statistic is the distribution of the values of the statistic calculated for an infinite number of datasets produced by the same Data Generating Process (DGP). The DGP includes all the relevant factors that affect the data, including not only characteristics of the population under study, but also characteristics of the study, such as whether participants were randomly sampled, how many participants were included, which measurement tools were used, etc. Choices such as when to stop collecting data are part of the study design. That means that the same data can have a different $p$-value, depending on whether the sample size was fixed a priori, or whether sampling continued until some criterion was reached. The following story, paraphrased from [@berger, p. 30–33], may highlight the issue:

A scientist has obtained 100 independent observations that are assumed be Normal-distributed with mean $\mu$ and standard deviation $\sigma$. In order to test the null hypothesis that $\mu=0$, the scientist consults a Frequentist statistician. The mean of the observations is $\overline{Y} = 0.2$, and the sample standard deviation is $S_Y=1$, hence the $p$-value is $p = `r pvalue(2*(1-pt((.2-0)/(1/sqrt(100)),df=99)))`$, which is a little lower than than the adopted significance level of $\alpha.05$. This leads to a rejection of the null hypothesis, and a happy scientist. However, the statistician decides to probe deeper and asks the scientist what he would have done in case that the experiment had not yielded a significant result after 100 observations. The scientist replies he would have collected another 100 observations. As such, the implicit sampling plan was not to collect $n=100$ observation and stop, but rather to first take 100 observations and check whether $p <.05$, and collect another 100 observations (resulting in $n=200$) if not. This is a so-called sequential testing procedure, and requires a different treatment than a fixed-sampling procedure. In controlling the Type 1 error of the procedure as a whole, one would need to consider the possible results after $n=100$ observations, but also after $n=200$ observations, which is possible, but not straightforward, as the results of after $n=100$ are dependent on the results after $n=100$ observations. But the clever statistician works it out and then convinces the scientist that the appropriate p-value for this sequential testing procedure is no longer significant. The puzzled and disappointed scientist leaves to collect another 100 observations. After lots of hard work, the scientist returns, and the statistician computes a $p$-value for the new data, which is now significant. Just to make sure the sampling plan is appropriately reflected in the calculation, the statistician asks what the scientist would have done if the result would not have been significant at this point. The scientist answers "This would depend on the status of my funding; If my grant is renewed, I would test another 100 observations. If my grant is not renewed, I would have had to stop the experiment. Not that this matters, of course, because the data were significant anyway”. The statistician then explains that the correct inference depends on the grant renewal; if the grant is not renewed, the sampling plan stands and no correction is necessary. But if the grant is renewed, the scientist could have collected more data, which calls for a further correction, similar to the first one. The annoyed scientist then leaves and resolves to never again share with the statistician her hypothetical research intentions.

What this story shows is that in considering infinite possible repetitions of a study, everything about the study that might lead to variations in the results should be taken into account. This includes a scientists' decisions made during each hypothetical replication of the study. As such, the interpretation of the data at hand (i.e., whether the hypothesis test is significant or not significant) depends on hypothetical decisions in situations that did not actually occur. If exactly the same data had been collected by a scientist who would have not have collected more observations, regardless of the outcome of the first test, then the result would have been judged significant. So the same data can provide different evidence. This does not mean the Frequentist NHST is inconsistent. The procedure "does what it says on the tin", namely providing a bound on the rate of Type 2 errors _in decisions_, when the null hypothesis is true. 


You cannot "peek at the data", try out various predictors, etc., and calculate a "standard" $p$-value!

### Results of a NHST are often misinterpreted

The $p$-value ($p$) is the probability of observing a particular value of a test statistic, or one more extreme, given that the null-hypothesis is true.
  
Common misconceptions:

1. $p = P(H_0|\text{data})$, the probability that the null-hypothesis is true, given the data
2. $1-p = P(H_1|\text{data})$, the probability that the alternative hypothesis is true, given the data
3. $p$ = the probability that the results were due to random chance

## To Bayes or not to Bayes? A pragmatic view








Suppose someone wakes up with a fever. He immediately thinks of one thing: COVID-19. What is the probability that this person actually has the disease? The most common signs and symptoms of COVID-19, with their estimated prevalence, are given in Table \@ref(tab:covid-symptoms-table}). 

```{r covid-symptoms-table}
tab <- data.frame(
  symptom = c("fever","dry cough","fatigue","sputum production", "shortness of breath", "sore throat", "headache", "myalgiaor arthralgia", "chills", "nausea or vomiting", "nasal congestion", "diarrhea", "hemoptysis", "conjunctival congestion"),
  percentage = c("87.9%","67.7%","38.1%", "33.4%", "18.6%", "13.9%", "13.6%", "14.8%", "11.4%", "5.0%", "4.8%", "3.7%", "0.9%", "0.8%"))
knitr::kable(tab, align = c("l","r"), caption = "Signs and symptom of COVID-19 in 55,924 laboratory confirmed cases. [Source](https://www.who.int/docs/default-source/coronaviruse/who-china-joint-mission-on-covid-19-final-report.pdf): Report of the WHO-China Joint Mission on Coronavirus Disease 2019 (COVID-19)")
```

You can see that fever is present in `r tab[1,2]` of confirmed cases. Does that mean that the probability that he has COVID-19 is .879? I hope you thought: No! The value in the table is the conditional probability $P(\text{fever}|\text{COVID-19})$. What we would like to know however is a different conditional probability, namely $P(\text{COVID-19}|\text{fever})$. And these are _not_ the same! 

But of course, there are other possibilities. Fever is also a common sign of seasonal influenza, and one study estimates the prevalence of this at 68% in 2470 confirmed cases [@monto_2000_influenza]. 