```{r, echo=FALSE}
set.seed(20221010)
```

# Latent variable models (SEM 2) {#ch-SEM-latent-variable-models}

The previous chapter concerned a multivariate model to describe relations between *observed* variables. In this Chapter, we will extend this idea by adding **latent variables**. Latent variables are variables which can not be directly observed. Instead, they can be measured or inferred via their relation with observed variables. A classic example in psychology is intelligence.

Another example is a personality trait such as extraversion. People's tendency to be outwards facing is not directly observable. However, there are indicators (e.g., someone actively seeking out busy social situations, liking to be the centre of attention, etc.) which together may allow one to determine a person's level of extraversion. 

The measurement of psychological constructs is the focus of **psychometrics**, a field with a long history and its own societies and journals. 


## Measurement of latent variables

Let's start with a simple example: measuring temperature. Temperature is a physical quantity that reflects the kinetic energy of atoms in a substance or the air. This kinetic energy is not directly observable, but can be measured with a thermometer. A classic mercury thermometer has a bulb filled with mercury attached to a narrow tube of glass. An increase in heat expands the volume of the mercury so that the level of mercury in the narrow tube rises. In other words, the temperature is the *cause* of the state of the measurement device. Markers along the narrow tube can be placed such that the level of mercury in the tube corresponds to a standard scale, such as the Fahrenheit or Celcius scale. In the latter, the scale is calibrated such that $0^\circ \textrm{C}$ represents the freezing point of water, and $100^\circ \textrm{C}$ represents the boiling point of water. In the Fahrenheit scale, the freezing point of water corresponds to $32^\circ \textrm{F}$, and the boiling point to $212^\circ \textrm{F}$. 

Thermometers are relatively accurate measurement devices, but not devoid of **measurement error**. That means that the measurement device can provide different readings for exactly the same true temperature. Such measurement error may be due to factors affecting the measurement device (e.g. the volume of mercury not being deterministically related to the temperature) as well as factors affecting the observer (e.g. the person reading the value of the thermometer not being able to see very small changes on the thermometer).

As a statistical model, we might propose something like the following linear model:
$$\texttt{measurement}_i = \beta_0 + \beta_1 \times \textrm{temperature}_i + \epsilon_i \quad \quad \epsilon_i \sim \mathbf{Normal}(0, \sigma_\epsilon)$$
If temperature was a directly observed variable, this is just a simple regression model and we would be able to estimate all parameters ($\beta_0$, $\beta_1$, $\sigma_\epsilon$) using e.g. maximum likelihood. However, temperature is a latent variable and we can only observe the measurement. 

```{r}
library(lavaan)
library(semPlot)
dat <- data.frame(x = rnorm(100)) |>
  dplyr::mutate(measurement = 30 + 10*x)
mod_spec <- '
temperature =~ measurement
'
mod <- lavaan::cfa(mod_spec, data=dat)
spmod <- semPlotModel(mod)
#spmmod@Pars <- spmmod@Pars[-5,]
semPaths(spmod, sizeMan=7, sizeInt = 4, style="mx", residuals=TRUE, normalize=FALSE, width=4, height=2, rotation=2, nCharNodes = 0)
```


### Scaling and identification

As we start working towards general SEM models with latent variables, let's write down our model for temperature measurements in more abstract terms as:
$$\begin{aligned}
Y_{i} &= \alpha + \lambda \times \eta_{i} + \epsilon_{i} \\
\eta_i &\sim \mathbf{Normal}(\mu_\eta, \sigma_\eta)\\
\epsilon_i &\sim \mathbf{Normal}(0, \sigma_\epsilon)
\end{aligned}$$
Note that in addition to relabelling the intercept as $\alpha$, the slope as $\lambda$, and the latent variable as $\eta$, we have also added an assumption about the values of the latent variable, namely that these are Normal-distributed. We need to make such an assumption about the distribution of the latent variable because without it, we cannot infer properties of the latent variable from measurements $Y$. 

The model above implies that $Y$ is Normal-distributed with
$$\begin{aligned}
\mu_Y &= \alpha + \lambda \times \mu_\eta \\
\sigma^2_Y &= \lambda^2 \times \sigma^2_\eta + \sigma^2_\epsilon
\end{aligned}
$$
From observations of $Y$, we can estimate the mean and variance of $Y$ in the usual manner as $\hat{\mu}_Y = \overline{Y}$ and $\hat{\sigma}^2_Y = \frac{\sum_{i=1}^n (Y_i - \overline{Y})^2}{n-1}$. But our latent variable model has 5 parameters. We cannot estimate 5 parameters from what are effectively two properties of the data. 

There are two common choices to scale latent variables. The first is to set $\mu_\eta = 0$ and $\lambda = 1$. The second is to set $\mu_\eta = 0$ and $\sigma^2_\eta = 1$. 

Now, suppose we have readings of *two* thermometers,

$$\begin{aligned}
Y_{1,i} &= \alpha_1 + \lambda_1 \times \eta_{i} + \epsilon_{1,i} \\
Y_{2,i} &= \alpha_2 + \lambda_2 \times \eta_{i} + \epsilon_{2,i} \\
\eta_i &\sim \mathbf{Normal}(0, \sigma_\eta)\\
\epsilon_{1,i} &\sim \mathbf{Normal}(0, \sigma_{\epsilon_1}) \\
\epsilon_{2,i} &\sim \mathbf{Normal}(0, \sigma_{\epsilon_2})
\end{aligned}$$

\begin{equation}
\begin{aligned}
\left( \begin{matrix} Y_{1,i} \\ Y_{2,i} \end{matrix} \right)
&\sim \mathbf{Normal}\left(\boldsymbol{\mu}_Y, \boldsymbol{\Sigma}_Y \right) \\
&\sim \mathbf{Normal}\left( \left[ \begin{matrix} \alpha_1 \\ \alpha_2 \end{matrix} \right] , \left[ \begin{matrix} \lambda_1^2 \times \sigma_{\eta}^2 + \sigma_{\epsilon_1}^2 & \lambda_1 \times \lambda_2 \times \sigma^2_\eta  \\  \lambda_1 \times \lambda_2 \times \sigma^2_\eta & \lambda_2^2 \times \sigma_{\eta}^2 + \sigma_{\epsilon_2}^2 \end{matrix} \right] \right)
\end{aligned}
\end{equation}

But if we have readings from *three* thermometers, we finally get somewhere.

$$\begin{aligned}
Y_{1,i} &= \alpha_1 + \lambda_1 \times \eta_{i} + \epsilon_{1,i} \\
Y_{2,i} &= \alpha_2 + \lambda_2 \times \eta_{i} + \epsilon_{2,i} \\
Y_{3,i} &= \alpha_3 + \lambda_3 \times \eta_{i} + \epsilon_{3,i} \\
\eta_i &\sim \mathbf{Normal}(0, \sigma_\eta)\\
\epsilon_{1,i} &\sim \mathbf{Normal}(0, \sigma_{\epsilon_1}) \\
\epsilon_{2,i} &\sim \mathbf{Normal}(0, \sigma_{\epsilon_2}) \\
\epsilon_{3,i} &\sim \mathbf{Normal}(0, \sigma_{\epsilon_3})
\end{aligned}$$
Which implies
\begin{equation}
\begin{aligned}
\left( \begin{matrix} Y_{1,i} \\ Y_{2,i} \\ Y_{3,i} \end{matrix} \right)
&\sim \mathbf{Normal}\left( \left[ \begin{matrix} \alpha_1 \\ \alpha_2 \\ \alpha_3 \end{matrix} \right] , \left[ \begin{matrix} \lambda_1^2 \times \sigma_{\eta}^2 + \sigma_{\epsilon_1}^2 & \lambda_1 \times \lambda_2 \times \sigma^2_\eta & \lambda_1 \times \lambda_3 \times \sigma^2_\eta \\  \lambda_1 \times \lambda_2 \times \sigma^2_\eta & \lambda_2^2 \times \sigma_{\eta}^2 + \sigma_{\epsilon_2}^2 & \lambda_2 \times \lambda_3 \times \sigma^2_\eta \\ \lambda_1 \times \lambda_3 \times \sigma^2_\eta & \lambda_2 \times \lambda_3 \times \sigma^2_\eta & \lambda_3^2 \times \sigma_{\eta}^2 + \sigma_{\epsilon_3}^2 \end{matrix} \right] \right)
\end{aligned}
\end{equation}

## Confirmatory factor analysis

"Big Five" personality traits such as extraversion are generally assessed with self-report questionnaires. One commonly used questionnaire is the Big Five Inventory [BFI; @john1991big]. In the BFI-2 [@@soto2017next], participants are asked to rate their agreement with 60 statements on a 5-point scale ranging "disagree strongly" to "agree strongly". There are 12 statements relating to each of the following main personality dimensions: Extraversion, Agreeableness, Conscientiousness, Negative Emotionality (traditionally called "neuroticism"), and Open-Mindedness (tranditionally called "openness to experience"). Example statements are related to extraversion are _"I am someone who is outgoing, sociable"_ and _"I am am someone who is sometimes shy, introverted"_. Note that agreement with the first statement should be *positively* related to extraversion, whilst agreement with the second statement should be *negatively* related to extraversion. Mixing such positively and negatively stated items is common practice in the design of psychological measurement questionnaires. 

In the following, we will consider data collected with a Czech translation of the BFI-2 [@hvrebivckova2020big], with a total of $n=1733$ respondents.

```{r SEM-estimate-BFI2-extraversion-models}
library(tidyverse)
data(BFI2, package = "ShinyItemAnalysis")
dat <- BFI2
colnames(dat)[1:60] <- c(
  "iEscb01", "iAcmp02", "iCorg03r", "iNanx04r", "iOaes05r", "iEasr06",
  "iArsp07", "iCprd08r", "iNdep09r", "iOint10", "iEenl11r", "iAtrs12r", "iCrsp13", "iNemt14",
  "iOcrt15", "iEscb16r", "iAcmp17r", "iCorg18", "iNanx19", "iOaes20", "iEasr21", "iArsp22r",
  "iCprd23r", "iNdep24r", "iOint25r", "iEenl26r", "iAtrs27", "iCrsp28r", "iNemt29r",
  "iOcrt30r", "iEscb31r", "iAcmp32", "iCorg33", "iNanx34", "iOaes35", "iEasr36r", "iArsp37r",
  "iCprd38", "iNdep39", "iOint40", "iEenl41", "iAtrs42r", "iCrsp43", "iNemt44r", "iOcrt45r",
  "iEscb46", "iAcmp47r", "iCorg48r", "iNanx49r", "iOaes50r", "iEasr51r", "iArsp52", "iCprd53",
  "iNdep54", "iOint55r", "iEenl56", "iAtrs57", "iCrsp58r", "iNemt59", "iOcrt60"
)
dat <- dat %>%
  mutate(across(ends_with("n"), ~ 6 - .x))
colnames(dat)[1:60] <- paste0("i",1:60)

library(lavaan)
mod_spec <- '
E =~ i1 + i16 + i31 + i46 + i6 + i21 + i36 + i51 + i11 + i26 + i41 + i56
'
BFI2_extr_1 <- lavaan::cfa(mod_spec, data=dat, orthogonal=TRUE)
mod_spec <- '
E =~ i1 + i16 + i31 + i46 + i6 + i21 + i36 + i51 + i11 + i26 + i41 + i56
A =~ 1*i1 + 1*i16 + 1*i31 + 1*i46 + 1*i6 + 1*i21 + 1*i36 + 1*i51 + 1*i11 + 1*i26 + 1*i41 + 1*i56
'
BFI2_extr_1acq <- lavaan::cfa(mod_spec, data=dat, orthogonal=TRUE)
mod_spec <- '
E =~ E1 + E2 + E3
E1 =~ 1*i1 + i16 + i31 + i46 
E2 =~ 1*i6 + i21 + i36 + i51 
E3 =~ i11 + i26 + 1*i41 + i56
E ~~ 1*E
'
BFI2_extr_31 <- lavaan::lavaan(mod_spec, data=dat, int.ov.free = TRUE, int.lv.free = FALSE, auto.fix.first = FALSE, auto.fix.single = TRUE, auto.var = TRUE, auto.cov.lv.x = TRUE, auto.efa = TRUE, auto.th = TRUE, auto.delta = TRUE, auto.cov.y = TRUE)
mod_spec <- '
E =~ E1 + E2 + E3
E1 =~ 1*i1 + i16 + i31 + i46 
E2 =~ 1*i6 + i21 + i36 + i51 
E3 =~ i11 + i26 + 1*i41 + i56
A =~ 1*i1 + 1*i16 + 1*i31 + 1*i46 + 1*i6 + 1*i21 + 1*i36 + 1*i51 + 1*i11 + 1*i26 + 1*i41 + 1*i56
E ~~ 0*A
E ~~ 1*E
'
BFI2_extr_31acq <- lavaan::lavaan(mod_spec, data=dat, int.ov.free = TRUE, int.lv.free = FALSE, auto.fix.first = FALSE, auto.fix.single = TRUE, auto.var = TRUE, auto.cov.lv.x = TRUE, auto.efa = TRUE, auto.th = TRUE, auto.delta = TRUE, auto.cov.y = TRUE)
```

### A single factor model

The BFI2 contains 12 items which relate to extraversion. As these items pertain to the same psychological dimension, we would expect answers to them to be correlated. A confirmatory factor model aims to account for these correlations by assuming that all answers are "caused" by a single underlying latent factor. 

```{r BFI2-extr-1-plot, fig.cap="Graphical representation of a single factor model for Extraversion. Items i1, i16, i31, and i46 relate to the Sociability facet of Extraversion, with items i16 and i31 stated in the negative direction. Items i6, i21, i36, and i51 relate to the Assertiveness facet of Extraversion, with items i36 and i51 stated in the negative direction. Items 11, 26, 41, 56 relate to the Energy Level facet of Extraversion, with items i11 and i26 stated in the negative direction.", fig.width=8, fig.height=3.5, out.width="100%"}
semPlot::semPaths(BFI2_extr_1, whatLabels = "est")
# 3, 4, 5, 8, 9, 11, 12, 16, 17, 22, 23, 24, 25, 26, 28, 29, 30, 31, 36, 37, 42, 44, 45, 47, 48, 49, 50, 51, 55, and 58
```

Figure \@ref(fig:BFI2-extr-1-plot2) shows the estimated parameters of a single factor model for the 12 items in the BFI-2 related to extraversion. The factor loading of the first item was fixed to 1 in order to scale the latent factor. Note that all estimated factor loadings are positive, whilst 6 of the 12 items were stated in a negative direction (i.e. higher scores should indicate *less* extraversion). If all items relate to extraversion in the intended manner, we should expect a mix of positive and negative factor loadings. Table \@ref(tab:BFI2-extr-1-results) shows the parameter estimates and test results, as well as the model fit indices. As can be seen there, the single factor model is rejected by the overall model $\chi^2$ test, and the CFI and RMSEA both indicate relatively poor model fit.

```{r, results="asis"}
if(knitr::is_html_output()) {
  cat("<table><caption>(\\#tab:BFI2-extr-1-results) Results of a single factor model for extraversion.</caption>")
  semTable::semTable(BFI2_extr_1, type="html", fits = c("chisq", "cfi", "srmr", "rmsea", "rmsea.ci.lower", "rmsea.ci.upper"), fitLabels = c(rmsea.ci.lower = "RMSEA (lower bound)", rmsea.ci.upper = "RMSEA (upper bound)"))
  cat("</table>")
} else {
  semTable::semTable(fmod_complex2, type="latex", table.float = TRUE, fits = c("chisq", "cfi", "srmr", "rmsea", "rmsea.ci.lower", "rmsea.ci.upper"), caption = "(\\#ttab:tab:sem_path_mod_complex_results2) Results of a second full path model for the legacy motive data.")
}
```

@soto2017next suggest that participants' responses to the BFI-2 might be affected by their "acquiescent response style". This refers to the tendency of an individual to
consistently agree or consistently disagree with questionnaire items, regardless of their content. This would distort the factor loadings. They suggest to account for Acquiescence by including a second factor in the model, with fixed factor loadings on all items. By setting the factor loadings as identical for all items, this factor can account for a general tendency to provide low or high scores on all items. 

```{r BFI2-extr-1acq-plot, fig.cap="Graphical representation of a single factor model for Extraversion with an additional Acquiescence factor.", fig.width=8, fig.height=3.5, out.width="100%"}
semPlot::semPaths(BFI2_extr_1acq, whatLabels = "est", layout="tree2")
# 3, 4, 5, 8, 9, 11, 12, 16, 17, 22, 23, 24, 25, 26, 28, 29, 30, 31, 36, 37, 42, 44, 45, 47, 48, 49, 50, 51, 55, and 58
```

By fixing all the factor loadings, this additional factor only requires a single additional parameter which reflects the variance of the latent Acquiescence factor. The resulting model is depicted in Figure \@ref(fig:BFI2-extr-1acq-plot) and the resulting estimates and fit measures are provided in Table \@ref(tab:BFI2-extr-1acq-results). Whilst there are now a few negative factor loadings and the fit is somewhat improved, the model is still not adequate. 

```{r, results="asis"}
if(knitr::is_html_output()) {
  cat("<table><caption>(\\#tab:BFI2-extr-1acq-results) Results of a single factor model for Extraversion with an additional Acquiescence factor.</caption>")
  semTable::semTable(BFI2_extr_1acq, type="html", fits = c("chisq", "cfi", "srmr", "rmsea", "rmsea.ci.lower", "rmsea.ci.upper"), fitLabels = c(rmsea.ci.lower = "RMSEA (lower bound)", rmsea.ci.upper = "RMSEA (upper bound)"))
  cat("</table>")
} else {
  semTable::semTable(fmod_complex2, type="latex", table.float = TRUE, fits = c("chisq", "cfi", "srmr", "rmsea", "rmsea.ci.lower", "rmsea.ci.upper"), caption = "(\\#ttab:tab:sem_path_mod_complex_results2) Results of a second full path model for the legacy motive data.")
}
```


### CFA with multiple latent 

```{r SEM-estimate-BFI2-full-factor-models}
data(BFI2, package = "ShinyItemAnalysis")
dat <- BFI2
library(lavaan)
mod_spec <- '
E =~ i1 + i16 + i31 + i46 + i6 + i21 + i36 + i51 + i11 + i26 + i41 + i56
A =~ i2 + i17 + i32 + i47 + i7 + i22 + i37 + i52 + i12 + i27 + i42 + i57
C =~ i3 + i18 + i33 + i48 + i8 + i23 + i38 + i53 + i13 + i28 + i43 + i58
N =~ i4 + i19 + i34 + i49 + i9 + i24 + i39 + i54 + i14 + i29 + i44 + i59
O =~ i10 + i25 + i40 + i55 + i5 + i20 + i35 + i50 + i15 + i30 + i45 + i60
'
BFI2_cfa <- lavaan::cfa(mod_spec, data=dat, orthogonal=TRUE)
BFI2_cfa_dep <- lavaan::cfa(mod_spec, data=dat)
```



$$\texttt{measurement} = \beta_0 + \beta_1 \times \textrm{temperature} + \epsilon \quad \quad \epsilon \sim \mathbf{Normal}(0, \sigma_\epsilon)$$
where the error term $\epsilon$ is the combination of all factors that lead to measurement error, and the constants $\beta_0$ and $\beta_1$ define the scale used (e.g. Celcius, or Fahrenheit). As the scale is -- to some extent -- arbitrary, we could assume a scale where $\beta_0 = 0$ and $\beta_1 = 1$, leading to the model
$$\textrm{temperature}  =  \frac{\texttt{measurement} - \beta_0 - \epsilon}{\beta_1} \quad \quad \epsilon \sim \mathbf{Normal}(0, \sigma_\epsilon)$$

## Factor analysis

```{r}
library(psychTools)
data(bfi)
```


## Modification indices

Modification indices 

## General SEMs with latent variables

The BFI2 was designed with a hierarchical structure in mind. Indeed, each of the "big five" factors was assumed to be further broken down into three facets: extraversion into sociability, assertiveness and energ level, Agreeableness into compassion, respectfulness, and trust, conscientiousness into organization, productiveness, and responsibility, neuroticism into anxiety, depression, and emotional volatility, and openness into intellectual curiosity, aesthetic sensitivity, and creative imagination. 


```{r}
mod_spec <- '
E =~ E1 + E2 + E3
E1 =~ i1 + i16 + i31 + i46 
E2 =~ i6 + i21 + i36 + i51 
E3 =~ i11 + i26 + i41 + i56

A =~ A1 + A2 + A3
A1 =~ i2 + i17 + i32 + i47
A2 =~ i7 + i22 + i37 + i52
A3 =~ i12 + i27 + i42 + i57

C =~ C1 + C2 + C3
C1 =~ i3 + i18 + i33 + i48
C2 =~ i8 + i23 + i38 + i53
C3 =~ i13 + i28 + i43 + i58

N =~ N1 + N2 + N3
N1 =~ i4 + i19 + i34 + i49
N2 =~ i9 + i24 + i39 + i54 
N3 =~ i14 + i29 + i44 + i59

O =~ O1 + O2 + O3
O1 =~ i10 + i25 + i40 + i55
O2 =~ i5 + i20 + i35 + i50 
O3 =~ i15 + i30 + i45 + i60
'
BFI2_cfa_hier <- lavaan::cfa(mod_spec, data=dat, orthogonal=TRUE)
```

```{r}
library(sdamr)
data(speeddate)
speed_cor <- cor(speeddate[,c("other_attr", "other_sinc", "other_intel", "other_fun", "other_amb")], use="pairwise.complete.obs")
#psych::fa(speed_cor, nfactors = 5, n.obs=nrow(speeddate), rotate="none")

psych::fa(speeddate[,c("other_attr", "other_sinc", "other_intel", "other_fun", "other_amb")], nfactors = 4, rotate="none")
```

### Measurement model

The relations between a latent variable and the observed variables are called a **measurement model**.

```{r}

```

### Structural (path) model

### Indentification

Hayward case (negative errors)

## Multiple groups and measurement invariance

## Exploratory factor and principal components analysis

```{r}
library(psychTools)
data(bfi)

bfi_cor <- cor(bfi[,-c(26:28)], use="pairwise.complete.obs")
bfi_pca <- prcomp(bfi_cor)
summary(bfi_pca)

bfi_pca <- psych::principal(bfi_cor)
summary(bfi_pca)
```

Although not really a part of Structural Equation Models, it is instructive to also briefly discuss **data reduction** techniques such as principal components analysis and exploratory factor analysis. These methods aim to describe a $P \times P$ covariance (or correlation) matrix by means of relatively small set of variables. Essentially, both methods rely on saturated models which can replicate the sample covariance matrix perfectly.

The differences between principal component analysis and exploratory factor analysis are somewhat subtle. **Principal components analysis** (PCA) can be seen as applying a **formative model**, where observed variables "cause" the latent variables (called a principal components in this context). **Principal components** (which we will denote as $\text{PC}_j$) are simply linear functions of the variables:
$$\begin{aligned}
\text{PC}_1 &= w_{1,1} \times X_{1} + w_{1,2} \times X_{2} + w_{1,m} \times \ldots X_{m} \\
\text{PC}_2 &= w_{2,1} \times X_{1} + w_{2,2} \times X_{2} + w_{2,m} \times \ldots X_{m} \\
& \vdots \\
\text{PC}_m &= w_{m,1} \times X_{1} + w_{m,2} \times X_{2} + w_{m,m} \times \ldots X_{m}
\end{aligned}$$
Note that the principal components $\text{PC}_j$ have no error terms. They don't need to be estimated, but rather are *computed* from the variables $X_j$. Also note that we use a total of $m$ principal components to model a total of $m$ variables. 

By contrast, **exploratory factor analysis** (PCA) concerns a **reflective model**, where the latent variables "cause" the observed variables. A EFA model can be written as:
$$\begin{aligned}
Y_1 &= \lambda_{1,1} \times F_{1} + \lambda_{1,2} \times F_{2} + \ldots + \lambda_{1,q} \times  F_{q} + \epsilon_{1} && \epsilon_{1} \sim \mathbf{Normal}(0, \sigma_{\epsilon_1}) \\
Y_2 &= \lambda_{2,1} \times F_{1} + \lambda_{2,2} \times F_{2} + \ldots + \lambda_{2,q} \times F_{q} + \epsilon_{2} && \epsilon_{2} \sim \mathbf{Normal}(0, \sigma_{\epsilon_1}) \\
& \vdots \\
Y_m &= \lambda_{m,1} \times F_{1} + \lambda_{m,2} \times F_{2} + \ldots + \lambda_{m,q} \times F_{q} + \epsilon_{m} && \epsilon_{m} \sim \mathbf{Normal}(0, \sigma_{\epsilon_m}) 
\end{aligned}
(\#eq:sem-2-efa-definition)$$
with the number of factors $q < m$. In a EFA, the observed variables $Y_j$ are considered conditionally independent, given the factors $F_k$. This is evident from the (independently distributed) error terms $\epsilon_j$. 

```{r}
library(lavaan)
# mod_spec <- '
# pc <- 1*E1 + 1*E2 + 1*E3 + 1*E4 + 1*E5
# E1 ~ 1
# E2 ~ 1
# E3 ~ 1
# E4 ~ 1
# E5 ~ 1
# #E1 ~~ E2 + E3 + E4 + E5
# #E2 ~~ E3 + E4 + E5
# #E3 ~~ E4 + E5
# #E4 ~~ E5
# '
# pca_lmod <- lavaan::sem(mod_spec, data=bfi)
# 
# ### PCA vs EFA
# 
# mod_spec <- '
# efa("block1")f1 =~ E1 + E2 + E3 + E4 + E5
# efa("block2")f2 =~ E1 + E2 + E3 + E4 + E5
# #E1 ~ 1
# #E2 ~ 1
# #E3 ~ 1
# #E4 ~ 1
# #E5 ~ 1
# #E1 ~~ E2 + E3 + E4 + E5
# #E2 ~~ E3 + E4 + E5
# #E3 ~~ E4 + E5
# #E4 ~~ E5
# '
# pca_lmod <- lavaan::sem(mod_spec, data=bfi)

### PCA vs EFA
```




```{r}
### Venn diagram for PCA/EFA
```


### Determining the number of factors

### Factor rotation

There are two main types of factor rotation: in **orthogonal rotation**, the factors are constrained to be independent of each other (i.e. there is no residual covariation between the factors). In **oblique rotation**, factors are allowed to have residual covariation. 

### Confirmatory factor analysis

Like EFA, confirmatory factor analysis (CFA) concerns a reflective model, where a factor causes the item values. But unlike PCA and EFA, CFA does not generally employ a saturated model. Items generally have paths from only a small set of the total number of factors. Th

```{r}
### bfi CFA
mod_spec <- '
Fa =~ A1 + A2 + A3 + A4 + A5 
Fc =~ C1 + C2 + C3 + C4 + C5
Fe =~ E1 + E2 + E3 + E4 + E5
Fn =~ N1 + N2 + N3 + N4 + N5
Fo =~ O1 + O2 + O3 + O4 + O5
'

bfi_cfa <- lavaan::cfa(mod_spec, data=bfi)
```

## Ordinal variables

## Conclusion



<!-- use PoliticalDemocracy data from lavaan. Code from https://www.lavaan.ugent.be/tutorial/sem.html -->

<!-- ## Multi-group SEM -->

## In practice