---
title: "Statistics, data analysis, and modeling"
author: "Maarten Speekenbrink"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::gitbook:
    lib_dir: "book_assets"
  bookdown::pdf_book:
    keep_tex: yes
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: mspeekenbrink/sdam-book
description: "Book/course notes"
---

# Preface {-}

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.align = "center")

knitr::opts_hooks$set(echo = function(options) {
  if (options$fig.width < options$fig.height) {
    options$fig.width <- options$fig.height
  }
  options
})

library(sdamr)
```

```{r myFunctions, echo=FALSE}
options(digits=3)
pretNum <- function(x) {
  prettyNum(x, big.mark=",")
}
```

This book concerns __statistics__, and more specifically __data analysis__ and __modeling__. It aims to be a self-contained source, requiring little background in mathematics or statistics. That said, it is not necessarily an "easy" book. It contains some mathematical notation and ideas. I do my best to properly explain how to a read and interpret these in both an intuitive and more formal way. I don't do this to torture you. Neither you with little mathematical background, who might have to read some explanations twice or more, nor you with more mathematical background, who might find my intuitive explanations tedious, superfluous, and utterly pointless, not you in the middle, who like me, who will agree with both (but hopefully not simultaneously). What I hope to do is to bring you some sense of the beauty of a formal system that stems from its simplicity and exactness, as well as some of the excitement of grasping it. I'm under no illusion that this is an easy task, for you or for me. But I hope you will try, like I did writing this book.

We will focus on building models of data in order to evaluate claims about how that data came to be what it is. Statistics is often taught with a "cookbook" approach. The "ingredients" are characteristics of the data (e.g. a variable is metric) and the goal of the analysis (e.g. compare the mean of two groups) and the cookbook provides the recipe (e.g., a two-sample t-test). While straightforward to teach, the limitations of this approach become quickly apparent when you need to analyse data from a less "standard" design. The cookbook can also become something of a straightjacket, when you are forced to change the design of a study in order to allow the subsequent analysis to be covered in the cookbook. Another limitation of the approach is that the different analyses and tests are often treated in isolation, which makes it more difficult to see the often strong connections between the models and statistical principles underlying them. By contrast, the modeling approach adopted here is much more flexible. The aim is to give you the confidence to, when given the ingredients, to design your own recipe. 

## Acknowledgements {-}

This book is inspired by lots of other statistics books I have read over the years. Some books deserve special mention. "Data analysis: A model comparison approach" [@judd2011data] is a book I have used for many years as required reading for the MSc level statistics course. It is extremely clear in its coverage of the General Linear Model, and concistent in its use of the model comparison approach we adopt here as well. "Doing Bayesian data analysis" [@kruschke2014doing] gave me the idea tp cover the basics of statistical modeling with a simple binomial example.

## Notation {-}

In a book with mathematics, notation is important. 

```{r greek-symbol-table}
greek_symbols <- c("alpha","beta","gamma","delta","epsilon","zeta","eta","theta","iota","kappa","lambda","mu","nu","xi","omicron","pi","rho","sigma","tau","upsilon","phi","chi","psi","omega")
knitr::kable(
  data.frame(
    symbol = paste0("$\\",greek_symbols,"$"),
    name = greek_symbols,
    usage = c("significance level","parameter (slope or intercept) in the General Linear Model","","error","","","a parameter in general","","","","mean","","","","","correlation","standard deviation","","","","","Chi-squared distribution","","")
  )
)
```

```{r}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
